Combining network-guided GWAS to discover susceptibility mechanisms for breast cancer

### 2019-11-16 manuscript/manuscript.tex
updated with fabienne's comments

\title{Combining network-guided GWAS to discover susceptibility mechanisms for breast cancer}
\author{Héctor Climente-González$^{1,2,3}$,Christine Lonjou$^{1,2,3}$,Fabienne Lesueur$^{1,2,3}$,GENESIS
  investigators,Nadine Andrieu$^{1,2,3}$,Chloé-Agathe Azencott$^{3,1,2}$}
\affiliation{$^{1}$Institut Curie, PSL Research University, F-75005 Paris, France; \\
  $^{2}$INSERM, U900, F-75005 Paris, France; \\
  $^{3}$MINES ParisTech, PSL Research University, CBIO-Centre for Computational Biology, F-75006 Paris, France; \\
}
\hypersetup{
 pdfauthor={Héctor Climente-González, Chloé-Agathe Azencott},
 pdftitle={Combining network-guided GWAS to discover susceptibility mechanisms for breast cancer},
 pdfkeywords={},
 pdfsubject={},
 pdfcreator={Emacs 26.2 (Org mode 9.2.3)}, 
 pdflang={English}}
\begin{document}

@@

Systems biology provides a comprehensive approach to biomarker discovery and biological hypothesis building. It does so by jointly considering the statistical association between a gene and a phenotype, obtained experimentally, and the biological context of each gene, represented as a network. In this work we study the utility of six network methods to discover new biomarkers in GWAS
data by searching subnetworks highly associated to a phenotype. We interrogate a familial breast cancer GWAS focused on BRCA1/2 negative French women. By trading statistical astringency for biological meaningfulness, most network methods get more compelling results than standard SNP- and gene-level analyses, recovering causal subnetworks tightly related to cancer susceptibility. We
perform an in-depth benchmarking of the methods with regards to size of the solution subnetwork, their utility as biomarkers, and the stability and the runtime of the methods. Interestingly, a combination solution subnetworks provided a concise subnetwork of 93 genes, enriched in known familial breast cancer susceptibility genes (BABAM1, BLM, CASP8, FGFR2, and TOX3, Fisher's exact test
P-value = $7.8 \times 10^{-5}$ ) and more central than average. Additionally, it includes subnetworks of mechanisms related to cancer, like protein folding (HSPA1A, HSPA1B, and HSPA1L) or mitocondrial ribosomes (MRPS30, MRPS31, MRPS18B). We also observed a general disregulation in the neighborhood of COPS5, a gene related to multiple hallmarks of cancer.

@@

In human health, genome-wide association studies (GWAS) aim at quantifying how single-nucleotide polymorphisms (SNPs) predispose to complex diseases, like diabetes or some forms of cancer \cite{bush_chapter_2012}. To that end, in a typical GWAS thousands of unrelated samples are genotyped: the cases, suffering the disease of interest, and the controls from the general population. Then, a per-SNP statistical association test is conducted (e.g. \(\chi\)\textsuperscript{2}). Those SNPs with a P-value lower than a conservative Bonferroni threshold are candidates to further studies in an independent cohort. Once the risk SNPs have been discovered, they can be used for diagnosis, guide the treatment, and deepen our understanding of the disease.

GWAS have successfully identified thousands of variants underlying many common diseases \cite{buniello_nhgri-ebi_2019}. However, the experimental setting also presents intrinsic challenges. Some of them stem from the high-dimensionality of the problem, as every GWAS to date studies more variants than samples are genotyped. This limits the statistical power of the experiment, as only variants with large and moderate effects can be detected. And it is particularly problematic since the prevailing view is that most genetic architectures involve many variants with small effects \cite{visscher_10_2017}. Additionally, to avoid false positives, a conservative multiple test correction is applied, typically Bonferroni. However, Bonferroni is known to be overly conservative when the statistical tests are correlated, as is the case in GWAS \cite{wang_statistical_2018}. Another open issue is the interpretation of the results, as the functional consequences of most common variants are not well understood. The fact that SNPs in non-coding regions often top the results of GWAS illustrates well this challenge. On top of that, recent large-sampled studies suggest that most of the genome contributes to a degree to any complex trait, in accordance with the infinitesimal model \cite{barton_infinitesimal_2017}. The omnigenic model \cite{boyle_expanded_2017} explains this by the dense functional inter-relatedness between genes, influencing each others' behavior, which allows alterations in most genes to impact the ``core'' genes involved in a disease. A comprehensive statistical framework which includes the structure of biological data might address the aforementioned issues.

@@

In this work, we analyze the effectiveness of these six methods to discover new biomarkers on GWAS data. We focus on the GENESIS dataset \cite{sinilnikova_genesis:_2016}, a homogeneous dataset of familial breast cancer in the French population. After following a classical GWAS approach, we use these network-based methods to recover additional familial breast cancer biomarkers. Some of them are known, while others are specific to this dataset. Lastly, we carry out a comparison of the solutions obtained by the different methods, and aggregate them to obtain a consensus network of predisposition to familial breast cancer.

@@

GENESIS is a GWAS study of familial breast cancer on the French population \cite{sinilnikova_genesis:_2016}. The index cases are patients with infiltrating mammary or ductal adenocarcinoma, who had sister with breast cancer, and negative for BRCA1 and BRCA2 mutations. The controls are unaffected colleagues and/or friends of the cases, born around the year of birth of the corresponding case (\textpm{} 3 years). We focused on the 2\,577 samples of European ancestry, of which 1\,279 are controls and 1\,298 are cases. The genotyping platform was iCOGS, a custom Illumina array designed to study genetic susceptibility of hormone-related cancers \cite{sakoda_turning_2013}. It contains 210\,918 SNPs previously associated with cancer susceptibility, survival, or other cancer-related traits; and in selected genes and pathways.

@@

We discarded SNPs with a minor allele frequency lower than 0.1\%, not in Hardy - Weinberg equilibrium (P-value \textless 0.001), and those with missing values on more than 10\% of the samples. A subset of 20 duplicated SNPs in FGFR2 were also removed. In addition, we removed the samples with more than 10\% missing genotypes, and an additional 28 samples with TODO. The final dataset included 1\,271 controls and 1\,280 cases, genotyped over 197\,083 SNPs.

@@

\subsection{High-score subnetwork search algorithms}
\label{sec:org6ced6c9}
\subsubsection{SNP and gene association}
\label{methods:node_score}
To measure association between a genotype and the phenotype, we performed a per-SNP 1df \(\chi\)\textsuperscript{2} allelic test using PLINK v1.90 \cite{chang_second-generation_2015}. Then, we used VEGAS2v2 to compute the gene-level association score \cite{mishra_vegas2:_2015} from the SNP P-values. In order to map SNPs to genes we used their overlap on the genome: all SNPs located within the boundaries of a gene, \textpm{} 50 kb, were mapped to that gene. To compute the gene association we used the 10\% of SNPs with lowest P-values. We used the 62\,193 genes described in GENCODE 31 \cite{frankish_gencode_2019}, although only 54\,612 could be mapped to at least one SNP. Out of those, we focused exclusively on the 32\,767 that could be mapped to an HGNC symbol. Out of the SNPs 197\,083 remaining after quality control, 164\,037 mapped to at least one of these genes.

@@

In addition, we use several matrices that describe different properties of a graph. The described matrices are square, and have as many rows and columns as nodes are in the network. The element $(i,j)$ represents a  selected relationship between $v_{i}$ and $v_{j}$. The \emph{adjacency matrix} $W_{G}$ contains a 1 when the corresponding nodes are connected through an edge, and 0 otherwise; the diagonal is zero. The \emph{degree matrix} $D_{G}$ is a diagonal matrix which contains the degree of the different nodes. Lastly, the \emph{Laplacian matrix} $L_{G}$ is defined as $L_{G} = D_{G} - W_{G}$.

\subsubsection{Methods used}
\label{methods:methods}
\begin{table}[htbp]
\caption{\label{tab:method_comparison}
Summary of the differences between the studied algorithms.}
\centering
\begin{tabular}{l|lllllll}
Method & Field & Nodes & Exhaustive & Solution & Components & Input & Scoring\\
\hline
dmGWAS & GWAS & Gene & No & - & 1 & Summary & -log\textsubscript{10}(P)\\
heinz & Omics & Gene & Yes & - & 1 & Summary & BUM\\
HotNet2 & Omics & Gene & Yes & Modular & \(\ge\) 1 & Summary & Local FDR\\
LEAN & Omics & Gene & Yes & Star & \(\ge\) 1 & Summary & -log\textsubscript{10}(P)\\
SConES & GWAS & SNP & Yes & Modular & \(\ge\) 1 & Genotypes & \(\chi\)\textsuperscript{2}\\
SigMod & GWAS & Gene & Yes & Modular & 1 & Summary & -log\textsubscript{10}(P)\\
\end{tabular}
\end{table}

@@

\begin{description}
\item[{dmGWAS}] dmGWAS searches the subgraph with the highest local density in low P-values \cite{jia_dmgwas:_2011}. To that end it searches candidate subnetwork solutions using a greedy procedure involving the following steps:

@@

\subsubsection{SNP networks}
\label{methods:snp_network}
SConES \cite{azencott_efficient_2013} is the only of the studied methods designed to handle SNP networks. As in gene networks, two SNPs are linked in a SNP network when there is evidence of shared functionality between two SNPs. The authors suggested three ways of building these networks: connecting the those SNPs consecutive in the genomic sequence (``GS network''); interconnecting all the SNPs mapped to the same gene, on top of GS (``GM network''); and interconnecting all SNPs mapped to two genes for which a protein-protein interaction exists (``GI network''). We focused on the GI network, as it is the network that fits better the scope of this article. However, at different stages of the manuscript we also used GS and GM. For the GM network, we used the mapping described in Section \ref{methods:node_score}. For the GI network, we used the PPI as described in Section \ref{methods:gene_network}. For all three networks the node score used is the association of the individual SNPs with the phenotype; specifically, we used the 1 d.f. \(\chi\)\textsuperscript{2}.

@@

\subsection{Evaluation of methods}
\label{sec:orgc968928}
\subsubsection{Classification accuracy of selected biomarkers}
\label{methods:comparison}
A desirable solution is one that is sparse, while offering a good predictive power on unseen samples. We evaluated the predicting power of the SNPs selected by the different methods through the performance of an L1-penalized logistic regression trained exclusively on those SNPs to predict the outcome (case/control). The L1 penalty helps to account for LD to reduce the number of SNPs included in the model (size of the active set), while improving the generalization of the classifier. The value of the regularization parameter, which controls both the magnitude and the sparsity of the coefficients, was set by cross-validation. To that end, we used the different network-methods on a random subset of 80\% of the samples. On this same subset we trained our classifier exclusively on the SNPs selected by a particular method. When the method retrieved a list of genes (all of them except SConES), all the SNPs mapped to any of those genes were used. Then we evaluated performance of the classifier on the remaining 20\% of the dataset. We repeated this procedure 5 times to estimate the average and the deviation of the different performance measures. The different performance measures we used where: size of the solution, size of the active set, specificity, and sensitivity. The size of the active set provides an estimate of a plausible, more sparse solution with a comparable predictive power to the original solution.

@@

\subsubsection{Biological relevance of the genes}
\label{methods:bcac}
An alternative way to validate the results is comparing our results to an external dataset. For that purpose, we recovered a list of 153 genes associated to familial breast cancer from DisGeNET \cite{pinero_disgenet:_2017}. Across this article we refer to these genes as \emph{familial breast cancer genes}.

Additionally, we used the summary statistics from the Breast Cancer Association Consortium (BCAC) \cite{michailidou_genome-wide_2015}. BCAC is one of the largest efforts in GWAS, genotyping over 120\,000 women of European ancestry. As opposed to GENESIS, samples were not selected based on family history, and hence is enriched in sporadic breast cancers. Another difference is that BCAC is a relatively heterogeneous study on a pan-European sample, while GENESIS is a homogeneous dataset focused on the French population. Despite these differences, there should be shared genetic architecture. On top of that, that overlap should become larger when the results are aggregated at the gene level. For that purpose, we computed the gene association as in Section \ref{methods:node_score}. iCOGS array was used for genotyping in BCAC \cite{sakoda_turning_2013}, the same array as for GENESIS \cite{sinilnikova_genesis:_2016}. Although imputed data is available, we used exclusively the SNPs available on GENESIS after quality control to make the results comparable.

@@

This work required developing computational pipelines for several GWAS analyses, such physically mapping SNPs to genes, computing gene scores, and performing six different network analyses. For each of those processes, a streamlined, project-agnostic pipeline with a clear interface was created. They are compiled in the following GitHub repository: \url{https://github.com/hclimente/gwas-tools}. The code that applies these pipelines to the GENESIS project, as well as the code that reproduces all the analyses in this article are available at \url{https://github.com/hclimente/genewa}. Although the GENESIS data is not publicly available, the published code should work on any other GWAS dataset. All the produced gene subnetworks were deposited on NDEx (\url{http://www.ndexbio.org}), under the UUID e9b0e22a-e9b0-11e9-bb65-0ac135e8bacf.

\section{Results}
\label{sec:orgc718f04}
\subsection{FGFR2 is strongly associated with familial breast cancer}
\label{results:conventional}
\begin{figure}[htbp]
\centering
\includegraphics[width=.9\linewidth]{./figures/figure_1.png}
\caption{\label{fig:snp_gene_manhattan}
Association in GENESIS. The red line represents the Bonferroni threshold. \textbf{(A)} SNP association, measured from the outcome of a 1df \(\chi\)\textsuperscript{2} allelic test. Significant SNPs that are within a coding gene, or within 50 kilobases of its boundaries, are annotated. The Bonferroni threshold is 2.54 \texttimes{} 10\textsuperscript{-7}. \textbf{(B)} Gene association, measured by P-value of VEGAS2v2 \cite{mishra_vegas2:_2015} using the 10\% of SNPs with the lowest P-values. The Bonferroni threshold is 1.53 \texttimes{} 10\textsuperscript{-6}.}
\end{figure}

We conducted association analyses in the GENESIS dataset at both the SNP and the gene levels (Section \ref{methods:node_score}). Two genomic regions have a P-value lower than the Bonferroni threshold in chromosomes 10 and 16 (Figure \ref{fig:snp_gene_manhattan}A). The former overlaps with gene FGFR2; the latter with CASC16, and it is located near the protein-coding gene TOX3. Variants in both FGFR2 and TOX3 were related to breast cancer susceptibility in other cohorts negative for BRCA1/2 \cite{rinella_genetic_2013}. Only the peak in chromosome 10 replicated in the gene-level analysis, with FGFR2 just above the threshold of significance (Figure \ref{fig:snp_gene_manhattan}B).

These results show the overlap in the genetic architecture of the disease between the French population and other cohorts, especially at the gene level. In addition, there are other SNPs whose P-values, although higher than the conventional threshold of significance, show a strong association with familial breast cancer. The most prominent of such regions are 3p24 and 8q24, both of which  which have been associated to breast cancer susceptibility in the past \cite{brisbin_meta-analysis_2011,search_newly_2009}. This motivates exploring network methods, which trade statistical significance for biological relevance.

\subsection{Network methods successfully identify genes linked to breast cancer}
\label{results:separate_networks}
\begin{table}[htbp]
  \begin{threeparttable}
\caption{\label{tab:gene_solutions}
Summary statistics on the results of multiple network methods on the gene-gene interaction network. The first row contains the summary statistics on the whole network.}
\centering
\begin{tabular}{lrrrrr}
Network & \# genes & \# edges & \mean{Betweenness} & \median{P}\textsubscript{gene} & \(\rho\)\textsubscript{consensus}\\
\hline
HINT HT & 13\,619 & 142\,541 & 16\,706 & 0.46 & 0.066\\
\hline
Consensus & 55 & 117 & 74\,062 & 0.0051 & 1\\
dmGWAS & 194 & 450 & 49\,115 & 0.19 & 0.41\\
heinz & 4 & 3 & 113\,633 & 0.0012 & 0.21\\
HotNet2 & 440 & 374 & 7\,739 & 0.048 & 0.31\\
LEAN & 0 & 0 & - & - & -\\
SConES GI & 0 (1) & 0 & - & - & -\\
SigMod & 142 & 249 & 92\,603 & 0.0083 & 0.73\\
\end{tabular}
\begin{tablenotes}
\emph{\# genes}: number of genes selected out of those that are part of the PPIN; in parentheses, the total number of genes. \mean{Betweenness}: mean betweenness of the selected genes in the PPIN. \median{P}\textsubscript{gene}: median P-value of the selected genes. \(\rho\)\textsubscript{consensus}: Pearson's correlation between the subnetwork and the consensus network.
\end{tablenotes}
\end{threeparttable}
\end{table}

>>>>

\title{Combining network-guided GWAS to discover susceptibility mechanisms for breast cancer}
\author{Héctor Climente-González$^{1,2,3}$,Christine Lonjou$^{1,2,3}$,Fabienne Lesueur$^{1,2,3}$,GENESIS
  investigators,Dominique Stoppa-Lyonnet$^{4,5,6}$,Nadine Andrieu$^{1,2,3}$,Chloé-Agathe Azencott$^{3,1,2}$}
\affiliation{$^{1}$Institut Curie, PSL Research University, F-75005 Paris, France; \\
  $^{2}$INSERM, U900, F-75005 Paris, France; \\
  $^{3}$MINES ParisTech, PSL Research University, CBIO-Centre for Computational Biology, F-75006 Paris, France;\\
  $^4$Service de Génétique, Institut Curie, F-75005 Paris, France;
  $^5$INSERM, U830, F-75005 Paris, France;
  $^6$Université Paris Descartes;

}
\hypersetup{
 pdfauthor={Héctor Climente-González, Chloé-Agathe Azencott},
 pdftitle={Combining network-guided GWAS to discover susceptibility mechanisms for breast cancer},
 pdfkeywords={},
 pdfsubject={},
 pdfcreator={Emacs 26.2 (Org mode 9.2.3)}, 
 pdflang={English}}
\begin{document}

@@

Systems biology provides a comprehensive approach to biomarker discovery and biological hypothesis building. It does so by jointly considering the statistical association between a gene and a phenotype, and the biological context of each gene, represented as a network. In this work we study the utility of six network methods to discover new biomarkers for breast cancer susceptibility by searching subnetworks highly associated to a phenotype. We interrogate a familial breast cancer genome-wide association study (GWAS) focused on \emph{BRCA1/2} negative French women. By trading statistical astringency for biological meaningfulness, most network methods get more compelling results than standard SNP- and gene-level analyses, recovering causal subnetworks tightly related to cancer susceptibility. We perform an in-depth benchmarking of the methods with regards to size of the solution subnetwork, their utility as biomarkers, and the stability and the runtime of the methods. Interestingly, a combination of solution subnetworks provided a concise subnetwork of 93 genes, enriched in known breast cancer susceptibility genes (\emph{BABAM1}, \emph{BLM}, \emph{CASP8}, \emph{FGFR2}, and \emph{TOX3}, Fisher's exact test P-value = $7.8 \times 10^{-5}$ ) and more central than average. Additionally, it includes subnetworks of mechanisms related to cancer, like protein folding (\emph{HSPA1A}, \emph{HSPA1B}, and \emph{HSPA1L}) or mitochondrial ribosomes (\emph{MRPS30}, \emph{MRPS31}, \emph{MRPS18B}). We also observed a general dysregulation in the neighborhood of \emph{COPS5}, a gene related to multiple hallmarks of cancer.

@@

In human health, genome-wide association studies (GWAS) aim at quantifying how single-nucleotide polymorphisms (SNPs) predispose to complex diseases, like diabetes or some forms of cancer \cite{bush_chapter_2012}. To that end, in a typical GWAS thousands of unrelated samples are genotyped: the cases, suffering the disease of interest, and the controls from the general population. Then, a per-SNP statistical association test is conducted (e.g. logistic regression). Those SNPs with a P-value lower than a conservative Bonferroni threshold are candidates to further studies in an independent cohort. Once the risk SNPs have been discovered, they can be used for risk assessment, and to deepen our understanding of the disease.

GWAS have successfully identified thousands of variants underlying many common diseases \cite{buniello_nhgri-ebi_2019}. However, the experimental setting also presents intrinsic challenges. Some of them stem from the high-dimensionality of the problem, as every GWAS to date studies more variants than samples are genotyped. This limits the statistical power of the experiment, as only variants with large and moderate effects can be detected. And it is particularly problematic since the prevailing view is that most genetic architectures involve many variants with small effects \cite{visscher_10_2017}. Additionally, to avoid false positives, a conservative multiple test correction is applied, typically Bonferroni. However, Bonferroni is known to be overly conservative when the statistical tests are correlated, as is the case in GWAS \cite{wang_statistical_2018}. Another open issue is the interpretation of the results, as the functional consequences of most common variants are not well understood. On top of that, recent large-sampled studies suggest that most of the genome contributes to a degree to any complex trait, in accordance with the infinitesimal model \cite{barton_infinitesimal_2017}. The omnigenic model \cite{boyle_expanded_2017} explains this by the dense functional inter-relatedness between genes, influencing each other's behavior, which allows alterations in most genes to impact the ``core'' genes involved in a disease. A comprehensive statistical framework which includes the structure of biological data might address the aforementioned issues.

@@

In this work, we analyze the effectiveness of these six methods to discover new biomarkers on GWAS data. We focus on the GENESIS dataset \cite{sinilnikova_genesis:_2016}, a study of familial breast cancer conducted in the French population. After following a classical GWAS approach, we use these network-based methods to recover additional familial breast cancer biomarkers. Some of them are known, while others are specific to this dataset. Lastly, we carry out a comparison of the solutions obtained by the different methods, and aggregate them to obtain a consensus network of predisposition to familial breast cancer.

@@

The GENE Sisters (GENESIS) study was designed to investigate risk factors for familial breast cancer in the French population \cite{sinilnikova_genesis:_2016}. Index cases are patients with infiltrating mammary or ductal adenocarcinoma, who had a sister with breast cancer, and who have been tested negative for BRCA1 and BRCA2 pathogenic variants. Controls are unaffected colleagues and/or friends of the cases, born around the year of birth of the corresponding case (\textpm{} 3 years). We focused on the 2\,577 samples of European ancestry, of which 1\,279 are controls and 1\,298 are cases. The genotyping was performed using the iCOGS array, a custom Illumina array designed to study genetic susceptibility of hormone-related cancers \cite{sakoda_turning_2013}. It contains 211\,155 SNPs, including SNPs putatively associated with breast, ovarian, and prostate cancers, SNPs associated with survival after diagnosis, and SNPs associated to other cancer-related traits, as well as functional candidate variants in selected genes and pathways.

@@

We discarded SNPs with a minor allele frequency lower than 0.1\%, those not in Hardy - Weinberg equilibrium in controls (P-value \textless 0.001), and those missing on more than 10\% of the samples. A subset of 20 duplicated SNPs in \emph{FGFR2} were also removed. In addition, we removed the samples with more than 10\% missing genotypes, and an additional 28 samples with TODO. The final dataset included 1\,271 controls and 1\,280 cases, genotyped over 197\,083 SNPs.

@@

\subsection{High-score subnetwork search algorithms}
\label{sec:org6ced6c9}
\subsubsection{SNP and gene association}
\label{methods:node_score}
To measure association between a genotype and the phenotype, we performed a per-SNP 1df \(\chi\)\textsuperscript{2} allelic test using PLINK v1.90 \cite{chang_second-generation_2015}. Then, we used VEGAS2v2 to compute the gene-level association score \cite{mishra_vegas2:_2015} from the SNP P-values. In order to map SNPs to genes we used their overlap on the genome: all SNPs located within the boundaries of a gene, \textpm 50 kb, were mapped to that gene. To compute the gene association we used the 10\% of SNPs linked to the gene with lowest P-values. We used the 62\,193 genes described in GENCODE 31 \cite{frankish_gencode_2019}, although only 54\,612 could be mapped to at least one SNP. Out of those, we focused exclusively on the 32\,767 that could be mapped to an HGNC symbol. Out of the SNPs 197\,083 remaining after quality control, 164\,037 mapped to at least one of these genes.

@@

In addition, we use several matrices that describe different properties of a graph. The described matrices are square, and have as many rows and columns as nodes are in the network. The element $(i,j)$ represents a  selected relationship between $v_i$ and $v_j$. The \emph{adjacency matrix} $W_G$ contains a 1 when the corresponding nodes are connected through an edge, and 0 otherwise; the diagonal is zero. The \emph{degree matrix} $D_G$ is a diagonal matrix which contains the degree of the different nodes. Lastly, the \emph{Laplacian matrix} $L_G$ is defined as $L_G = D_G - W_G$.

\subsubsection{Methods used}
\label{methods:methods}

\begin{table}[htbp]
  \begin{threeparttable}
    \caption{\label{tab:method_comparison}
      Summary of the differences between the studied algorithms.}
    \centering
    \begin{tabular}{l|lllllll}
      Method & Field & Nodes & Exhaustive & Solution & Components & Input & Scoring & Reference\\
      \hline
      dmGWAS & GWAS & Genes & No & - & 1 & Summary & -log\textsubscript{10}(P) & \cite{jia_dmgwas:_2011}\\
      heinz & Omics & Genes & Yes & - & 1 & Summary & BUM & \cite{dittrich_identifying_2008}\\
      HotNet2 & Omics & Genes & Yes & Modular & \(\ge\) 1 & Summary & Local FDR & \cite{leiserson_pan-cancer_2015}\\
      LEAN & Omics & Genes & Yes & Star & \(\ge\) 1 & Summary & -log\textsubscript{10}(P) & \cite{gwinner_network-based_2016}\\
      SConES & GWAS & SNPs & Yes & Modular & \(\ge\) 1 & Genotypes & \(\chi\)\textsuperscript{2} & \cite{azencott_efficient_2013}\\
      SigMod & GWAS & Genes & Yes & Modular & 1 & Summary & -log\textsubscript{10}(P) & \cite{liu_sigmod:_2017}\\
    \end{tabular}
    \begin{tablenotes}
      /Field/: field in which the algorithm was developed. /Nodes/, the type of network, either gene (protein-protein interaction network usually) or a SNP network. /Exhaustive/: whether all the possible solutions given the selected hyperparameters are explored. /Solution/: additional properties are enforced on the solution subnetwork, other than being dense in high scores and connected. /Components/: number of connected subnetworks in the solution. /Input/: genotype data or GWAS summary statistics. /Scoring/: how SNP/gene P-values are transformed into node scores.
    \end{tablenotes}
  \end{threeparttable}
\end{table}

@@

\begin{description}
\item[{dmGWAS}] dmGWAS searches the subgraph with the highest local density in low P-values \cite{jia_dmgwas:_2011}. To that end it searches candidate subnetwork solutions using a greedy, ``seed and extend'', heuristic:

@@

\subsubsection{SNP networks}
\label{methods:snp_network}
SConES \cite{azencott_efficient_2013} is the only of the studied methods designed to handle SNP networks. As in gene networks, two SNPs are linked in a SNP network when there is evidence of shared functionality between two SNPs. The authors suggested three ways of building these networks: connecting the SNPs consecutive in the genomic sequence (``GS network''); interconnecting all the SNPs mapped to the same gene, on top of GS (``GM network''); and interconnecting all SNPs mapped to two genes for which a protein-protein interaction exists (``GI network''). We focused on the GI network, as it is the network that fits better the scope of this article. However, at different stages of the manuscript we also used GS and GM. For the GM network, we used the mapping described in Section \ref{methods:node_score}. For the GI network, we used the PPI as described in Section \ref{methods:gene_network}. For all three networks the node score used is the association of the individual SNPs with the phenotype; specifically, we used the 1 d.f. \(\chi\)\textsuperscript{2}.

@@

\subsection{Evaluation of methods}
\label{sec:orgc968928}
\subsubsection{Classification accuracy of selected biomarkers}
\label{methods:comparison}
A desirable solution is one that is sparse, while offering a good predictive power on unseen samples. We evaluated the predicting power of the SNPs selected by the different methods through the performance of an L1-penalized logistic regression trained exclusively on those SNPs to predict the outcome (case/control). The L1 penalty helps to account for LD to reduce the number of SNPs included in the model (size of the active set), while improving the generalization of the classifier. The value of the regularization parameter, which controls both the magnitude and the sparsity of the coefficients, was set by cross-validation. To that end, we used the different network-methods on a random subset of 80\% of the samples. On this same subset we trained our classifier exclusively on the SNPs selected by a particular method. When the method retrieved a list of genes (all of them except SConES), we considered as selected all the SNPs mapped to any of those genes were used. Then we evaluated performance of the classifier on the remaining 20\% of the dataset. We repeated this procedure 5 times to estimate the average and the deviation of the different performance measures. The different performance measures we used were: size of the solution, size of the active set, specificity, and sensitivity. The size of the active set provides an estimate of a plausible, more sparse solution with a comparable predictive power to the original solution.

@@

\subsubsection{Biological relevance of the genes}
\label{methods:bcac}
An alternative way to validate the results is comparing our results to an external dataset. For that purpose, we recovered a list of 153 genes associated to familial breast cancer from DisGeNET \cite{pinero_disgenet:_2017}. Across this article we refer to these genes as \emph{breast cancer susceptibility genes}.

Additionally, we used the summary statistics from the Breast Cancer Association Consortium (BCAC) \cite{michailidou_genome-wide_2015}. BCAC has conducted one of the largest efforts in GWAS, involving over 120\,000 women of European ancestry. As opposed to GENESIS, samples were not selected based on family history, and hence is enriched in sporadic breast cancers. Another difference is that BCAC is a relatively heterogeneous study on a pan-European sample, while GENESIS is a homogeneous dataset focused on the French population. Despite these differences, there should be shared genetic architecture. On top of that, that overlap should become larger when the results are aggregated at the gene level. For that purpose, we computed the gene association as in Section \ref{methods:node_score}. iCOGS array was used for genotyping in BCAC \cite{sakoda_turning_2013}, the same array as for GENESIS \cite{sinilnikova_genesis:_2016}. Although imputed data is available, we used exclusively the SNPs available on GENESIS after quality control to make the results comparable.

@@

This work required developing computational pipelines for several GWAS analyses, such physically mapping SNPs to genes, computing gene scores, and performing six different network analyses. For each of those processes, a streamlined, project-agnostic pipeline with a clear interface was created. They are compiled in the following GitHub repository: \url{https://github.com/hclimente/gwas-tools}. The code that applies these pipelines to the GENESIS project, as well as the code that reproduces all the analyses in this article are available at \url{https://github.com/hclimente/genewa}. Although the GENESIS dataset is not publicly available, the published code should work on any other GWAS dataset. All the produced gene subnetworks were deposited on NDEx (\url{http://www.ndexbio.org}), under the UUID e9b0e22a-e9b0-11e9-bb65-0ac135e8bacf.

\section{Results}
\label{sec:orgc718f04}
\subsection{A conventional GWAS shows that FGFR2 is strongly associated with familial breast cancer}
\label{results:conventional}
\begin{figure}[htbp]
\centering
\includegraphics[width=.9\linewidth]{./figures/figure_1.png}
\caption{\label{fig:snp_gene_manhattan}
Association in GENESIS. The red line represents the Bonferroni threshold. \textbf{(A)} SNP association, measured from the outcome of a 1df \(\chi\)\textsuperscript{2} allelic test. Significant SNPs that are within a coding gene, or within 50 kilobases of its boundaries, are annotated. The Bonferroni threshold is 2.54 \texttimes{} 10\textsuperscript{-7}. \textbf{(B)} Gene association, measured by P-value of VEGAS2v2 \cite{mishra_vegas2:_2015} using the 10\% of SNPs with the lowest P-values. The Bonferroni threshold is 1.53 \texttimes{} 10\textsuperscript{-6}.}
\end{figure}

We conducted association analyses in the GENESIS dataset at both the SNP and the gene levels (Section \ref{methods:node_score}). Two genomic regions have a P-value lower than the Bonferroni threshold in chromosomes 10 and 16 (Figure \ref{fig:snp_gene_manhattan}A). The former overlaps with gene \emph{FGFR2}; the latter with \emph{CASC16}, and it is located near the protein-coding gene \emph{TOX3}. Variants in both \emph{FGFR2} and \emph{TOX3} were related to breast cancer susceptibility in other cohorts negative for \emph{BRCA1/2} \cite{rinella_genetic_2013}. Only the peak in chromosome 10 replicated in the gene-level analysis, with \emph{FGFR2} just above the threshold of significance (Figure \ref{fig:snp_gene_manhattan}B).

These results show the overlap in the genetic architecture of the disease between the studied French population sample and other populations, especially at the gene level. In addition, there are other SNPs whose P-values, although higher than the conventional threshold of significance, show a strong association with familial breast cancer. The most prominent of such regions are 3p24 and 8q24, both of which have been associated to breast cancer susceptibility in the past \cite{brisbin_meta-analysis_2011,search_newly_2009}. This motivates exploring network methods, which trade statistical significance for biological relevance.

\subsection{Network methods successfully identify genes associated with breast cancer}
\label{results:separate_networks}
\begin{table}[htbp]
  \begin{threeparttable}
\caption{\label{tab:gene_solutions}
Summary statistics on the results of multiple network methods on the gene-gene interaction network. The first row contains the summary statistics on the whole network.}
\centering
\begin{tabular}{lrrrrr}
Network & \# genes & \# edges & \mean{Betweenness} & \median{P}\textsubscript{gene} & \(\rho\)\textsubscript{consensus}\\
\hline
HINT HT & 13\,619 & 142\,541 & 16\,706 & 0.46 & 0.066\\
\hline
Consensus & 55 & 117 & 74\,062 & 0.0051 & 1\\
dmGWAS & 194 & 450 & 49\,115 & 0.19 & 0.41\\
heinz & 4 & 3 & 113\,633 & 0.0012 & 0.21\\
HotNet2 & 440 & 374 & 7\,739 & 0.048 & 0.31\\
LEAN & 0 & 0 & - & - & -\\
SConES GI & 0 (1) & 0 & - & - & -\\
SigMod & 142 & 249 & 92\,603 & 0.0083 & 0.73\\
\end{tabular}
\begin{tablenotes}
\emph{\# genes}: number of genes selected out of those that are part of the PPIN; for SConES GI the total number of genes, including RNA genes, was added in parentheses. \mean{Betweenness}: mean betweenness of the selected genes in the PPIN. \median{P}\textsubscript{gene}: median P-value of the selected genes. \(\rho\)\textsubscript{consensus}: Pearson's correlation between the subnetwork and the consensus network.
\end{tablenotes}
\end{threeparttable}
\end{table}

We applied six network methods to the GENESIS dataset (Section \ref{methods:methods}), obtaining six solutions (Supplementary Figure \ref{sfig:overview_solutions}, Supplementary Files 1 and 2): one for each of the five gene-based methods (Section \ref{methods:gene_network}), and one for SConES GI (Section \ref{methods:snp_network}). The solutions are very heterogeneous (Table \ref{tab:gene_solutions} and Supplementary table \ref{tab:snp_solutions}): none of the subnetworks examined by LEAN is significant (adjusted P-value < 0.05), while HotNet2 produced the largest solution subnetwork with 440 genes. SConES GI failed to recover genes in the PPIN, but it recovered one genomic region mapped to RNA gene RNU6-420P. All solution subnetworks except LEAN's are, on average, more strongly associated to breast cancer than the whole PPIN (median P-values $\ll 0.46$), despite containing genes with higher P-values (Supplementary Figure \ref{sfig:manhattan_solutions}). This exemplifies the trade-off between statistical significance and biological relevance. However, there are nuances between solutions: heinz strongly favored highly associated genes, while dmGWAS is less conservative (median gene P-values 0.0012 and 0.19, respectively); SConES tended to select whole LD-blocks; and HotNet2 and SigMod were less likely to select lowly associated genes.
